Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-08-29 bmc_template/manuscript.tex
not done, continue on line 253 in the am

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous 
HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers 
with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model 
within each type of transition complexity \bmb{clarify?} ($\dbic$ = BIC - min(BIC)) assesses the overall 
goodness of fit of each model type. (2) Comparing average step-length by hour of day for the 
observed data and for data simulated from the models shows how well a particular class of 
models can capture diurnal variation in behaviour. (3) Comparing temporal autocorrelations
for the observed data and for data simulated from the models shows how well a particular class
of models can capture serial correlation at both short and long time scales.

@@

As the number of states increases, the number of free parameters in (homogeneous or heterogenous) FMMs and time-homogeneous HMMs will increase linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically (Eq.~\ref{eq:numparam}). When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters. \bmb{what does last sentence mean?? delete?}

We used simulations to predict expected hourly step lengths and
autocorrelation functions (ACF).  While the computation of expected
step length and ACF is straightforward for FMMs, and feasible for
homogeneous HMMs, the interaction between the geometric dwell time
within each state and the temporally varying interaction probabilities
makes it infeasible for more complex models.  We used this approach to
validate our models, comparing our simulated predictions with the
observed movements.  The more usual approach, generating predictions
from the expected step lengths conditional on the most likely state
sequence predicted by the Viterbi algorithm \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
because the states predicted by the Viterbi algorithm already rely on the observed data. This approach is useful to predict
missing data in the observation sequence, but because it is
conditional on the observed values, it can not reliably evaluate
goodness of fit for the different structural complexities of HMM
models.
\bmb{we might need to be more careful here. Apparently Zucchini/Langrock et al also define ``pseudo-residuals'';
can you look these up and see if that approach suffers from the same issues as Viterbi ... ?}

@@

%% fig 1 goes somewhere here

\bmb{first sentence belongs in ``Methods'', with a few more words of motivation/description. I know that makes the Results for this part very short, but that's OK}

We simulated 100 realizations of a two-state HMM with sinusoidal temporal transitions and fitted it with 2- to 5-state HMMs with and without temporal heterogeneity in the transition probabilities. Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).

@@

\bmb{doesn't this para. belong in methods??}
As a complement, we also fitted FMM and FMM with priors on state occupancy that varied sinusoidally over time to compare the temporal effects in goodness of fit (dashed lines).
% \todo{I took out ``hourly HMM'' here - more coherent to describe it above} 
As a reminder, FMMs assume that the latent state in each time step is 
\emph{independent} of the latent state at the previous time step; 
time-varying FMMs can accurately describe movement when behaviour can
change on a short time scale, but the average propensity for different
behaviours changes over time.

>>>>

A simple simulation was conducted to to see whether fitting homogeneous transition probability HMMs to heterogeneous transitions simulated data will over estimate the number of states and estimate the correct number of states using the heterogeneous transitioning model. We used 100 realizations of a two-state HMM with sinusoidal temporal transitions and fitted it with 2- to 5-state HMMs with and without temporal heterogeneity in the transition probabilities.

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous 
HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers 
with the most data ($> 9000 $ observations). (1) BIC was used to compare the goodness of fit of each model type. The model with the lowest BIC was selected to be the optimal-BIC model and all BICs were adjusted to $\dbic$ based on the optimal-BIC model ($\dbic$ = BIC - min(BIC)) . (2) Comparing average step-length by hour of day for the 
observed data and for data simulated from the models shows how well a particular class of 
models can capture diurnal variation in behaviour. (3) Comparing temporal autocorrelations
for the observed data and for data simulated from the models shows how well a particular class
of models can capture serial correlation at both short and long time scales. As a complement, we also fitted FMM and FMM with priors on state occupancy that varied sinusoidally over time to compare the temporal effects in goodness of fit As a reminder, FMMs assume that the latent state in each time step is \emph{independent} of the latent state at the previous time step; time-varying FMMs can accurately describe movement when behaviour can
change on a short time scale, but the average propensity for different
behaviours changes over time.

@@

As the number of states increases, the number of free parameters in (homogeneous or heterogenous) FMMs and time-homogeneous HMMs will increase linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically (Eq.~\ref{eq:numparam}). When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

We used simulations to predict expected hourly step lengths and
autocorrelation functions (ACF).  While the computation of expected
step length and ACF is straightforward for FMMs, and feasible for
homogeneous HMMs, the interaction between the geometric dwell time
within each state and the temporally varying interaction probabilities
makes it infeasible for more complex models.  We used this approach to
validate our models, comparing our simulated predictions with the
observed movements.  The more usual approach, generating predictions
from the expected step lengths conditional on the most likely state
sequence predicted by the Viterbi algorithm or pseduo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
because the predictions by these methods rely on the observed data. These approach is useful to predict
missing data in the observation sequence, but because it is
conditional on the observed values, it can not reliably evaluate
goodness of fit for the different structural complexities of HMM
models.

@@

%% fig 1 goes somewhere here
The simulation agrees with our hypothesis that homogeneous transition HMMs can over estimate the number of hidden states. Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).

@@
